%%% Lecture 4

\lecture[We finish the proof of the HAT. My first non-trivial homotopy groups.]{2021-10-25}
$(k\geq2)$ Set $u_i=f_i(\text{center of } I^n)\in I^n$. Assume without loss of generality that the first coordinates of $\nn{u}{k}$ are not all equal (if some of them are, we can "wiggle" $f_1$).

\input{Pictures/Lec4Pic1}

Choose $t\in (0,1)$ such that
\begin{itemize}
    \item $t$ is different from the first coordinates of $\nn{u}{k}$,
    \item for some $1\leq i\leq k$, the first coordinate of $u_i$ is smaller than $t$,
    \item for some $1\leq i\leq k$, the first coordinate of $u_i$ is larger than $t$.
\end{itemize}

Choose neighborhoods $\Uu_i$ of $u_i$ inside $f_i(\ring I^n)$ that do not intersect $\cb{t}\times I^\ni$, that is such that $\Uu_i$ lies "on the same side respect to $t$" as $u_i$.

Choose subcubes $Q_i$ inside $\ring I^n$ that contain the center and lie in in $f^{-1}(\Uu_i)$.

Last time we proved that $g$ is pair-homotopic to some $g':\pair\to\pairs$ such that
\[g'(I^n\sm\bigcup_{i=1,\dots,k} f_i(Q_i))\subset A\quad\text{and}\quad[g\circ f_i]=[g'\circ f_i]\text{ in }\pi_n\pairs^\#.\]

We precompose each $f_i:I^n\to I^n$ with the linear shrinking homotopy relative $Q_i$. Set $f'_i=f_i\circ \text{ end of shrinking}$. Then $g'\circ f_i$ is pair homotopic to $g'\circ f'_i$ and $f'_i(I^n)\subset\Uu_i$.

By replacing $g$ by $g'$ and $f_i$ by $f'_i$ we can therefore assume without loss of generality that $f_i(\ring I^n)$ lies on one side of $\cb{t}\times I^\ni$.

\input{Pictures/Lec4Pic2}\smallskip

Write $g=g_1+_t g_2$ by "cutting along $\cb{x_1=t}$".

Formally, $g_1(\xs)=g(t\xs)$ and $g_2=g((1-t)x_1+t,\dots,x_n)$.

Set $I_1=\cb{i\in I\mid u_i\text{ lies left of }\cb{x_1=t}}$ and $I_2=\cb{i\in I\mid u_i\text{ lies right of }\cb{x_1=t}}$, where $I=\cb{1,\dots,k}$.

Then by the inductive hypothesis:
\[[g]=[g_1]+[g_2]=\sum_{i\in I_1}\deg(f_i)[g_1\circ f_i]+\sum_{i\in I_2}\deg(f_i)[g_2\circ f_i]=\sum_{i\in I}\deg(f_i)[g\circ f_i].\]\qed

\section{My First Non-Trivial Homotopy Group}

\begin{theorem}
Let $n\geq2$ and assume the HAT in dimension $n$ as well as $\pi_{n-1}(S^{n-1})\cong\Z$. Then $\pi_n(S^n,*)$ is infinite cyclic.
\end{theorem}

\begin{proof}
Choose some point $z\in S^n$. Set $U=S^n\sm\cb{-z}$. Then \[\pi_n(S^n,z)=\pi_n(S^n,\cb{z},z)\underset{U\cong *}{\cong}\pi_n(S^n,U,z)\underset{\pi_1(U,z)=\cb{1}}{\cong}\pi_n(S^n,U,z)^\dagger\cong\pi_n(S^n,U)^\#\]
So we may show that $\pi_n(S^n,U)^\#\cong\Z$.

We show that $\pi_n(S^n,U)^\#$ is generated by the class of any pair map $\psi:\pair\to(S^n,U)$ such that $\psi(\de I^n)=\cb{z}$ and $\psi$ factors out a homeomorphism $I^n/\de I^n\cong S^n$.

Let $f:(I^n,\de I^n)\to(S^n,U)$ be any pair map. Set $V=S^n\sm\cb{z}$ so that $S^n=U\cup V$ is an open cover.

The Lebesgue Number lemma provides an $m\geq1$ so that each subcube of $I^n$ of side length $1/m$ is mapped by $f$ into $U$ or into $V$. Decompose $I^n$ into $m^n$ subcubes of side length $1/m$.

We define subspaces of $I^n$ in this way:
\begin{itemize}[label=-]
    \item $A_{-1}=\de I^n$
    \item $A_0=A_{-1}\cup\text{ vertices}$
    \item $A_1=A_0\cup\text{ edges}$
    \item $A_2=\cdots$
    \item $A_n=I^n$
\end{itemize}

We want to "improve" $f$ successively by pair homotopies to maps $f=f_{-1},f_0,f_1,\dots,f_{n-1}$ such that:
\begin{itemize}
    \item each $f_j$ is homotopic to $f$ relative $\de I^n$,
    \item each $f_j$ is admissible, i.e. it sends every subcube of side length $1/m$ to $U$ or to $V$,
    \item $f_j(A_j)\subset U$ for $j=-1,0,1,\dots,n-1$.
\end{itemize}

We proceed by induction on $j$. There is nothing to show for $j=-1$. Let now $j\geq 0$. We first modify $f_{j-1}$ and the faces of the $j$-cube. If such a face $Q$ is "good", i.e. sent by $f_{j-1}$ into $U$, we do not do anything to $Q$. Otherwise the $(j-1)$-subcube is mapped to $V$ and the restriction of $f_{j-1}$ to it is a pair map $(Q,\de Q)\to (V,V\cap U)$.

Claim. For $j<n$, any pair map $(I^j,\de I^j)\to (V,U\cap V)$ is homotopic relative $\de I^j$ to a map with image in $U\cap V$.

\begin{claimproof}
By stereographic projection $(V,U\cap V)$ is pair homotopic to $(\R^n,\R^n\sm\cb{0})$, i.e. we can costruct a pair map $g:(I^j,\de I^j)\to (\R^n,\R^n\sm\cb{0})$. Because $\de I^j$ is compact and $0\notin f(\de I^j)$, there is an $\epsilon>0$ such that $g(\de I^j)\cap(\epsilon\text{-ball around }0)=\emptyset$.

So $g:(I^j,\de I^j)\to(\R^n,\R^n\sm\ring B(\epsilon,0))$. Now, $\R^n$ can be obtained from $\R^n\sm\ring B(\epsilon,0)$ by attaching an $n$-cell. The cellular approximation theorem and the fact that $I^j$ is a $j$-dimensional CW-complex gives us a relative homotopy from $g$ to a cellular map. Since $j<n$, the cellular map has image in $\R^n\sm\ring B(\epsilon,0)\subset\R^n\sm\cb{0}$.
\end{claimproof}

We can now change $f_{j-1}|A_j$ into $f_j|A_j$ by a homotopy relative $A_{j-1}$ into a map that sends all $j$-cells to $U$.

We use the HEP for $(I^n,A_j)$ to extend $f_j$ to all of $I^n$; we use the HEP with target $U$ or with target $V$ to ensure that the map $f_j$ is again admissible.

After this inductive construction we can replace $f$ by $f_{n-1}$ and we have arranged without loss of generality that $f(A_{n-1})\subset U$.

We can now assume that $g:\pair\to(S^n,U)$ satisfies $g(A_{n-1})\subset U$ and each top-dimensional subcube is mapped to $U$ or to $V$.

We apply the HAT to this map $g$ with $\nn{f}{k}$ the reparametrization of those subcubes that are \textit{not} mapped into $U$ (and hence into $V$).

Then by the HAT we have:
\[[g]=\sum_i\pm[g\circ f_i]\text{ in }\pi_n(S^n,U)^\#\cong\pi_n(S^n,\cb{z},z)^\dagger\cong\pi_n(S,z)\]

We have gained that each summand on the right hand side is in the image of the homomorphism $\pi_n(V,V\cap U)^\#\to\pi_n(S^n,U)^\#$, which is then surjective. By the long exact homotopy group sequence of the pair $(V,V\cap U)$, we obtain:
\[\pi_n(V,V\cap U,z)=\pi_n(\R^n,\R^n\sm\cb{0},z)\cong\pi_\ni(\R^{n-1}\sm\cb{0},z)\cong\pi_\ni(S^\ni,z)\cong\Z\]
so $\pi_n(S^n,U)^\#$ is infinite cyclic.
\end{proof}

\section{Reminder on Simplicial Sets}\label{section:reminder-on-sset}

We denote by $\Delta$ the category with objects the sets $[n]=\cb{0,1,\dots,n}$, $n\geq 0$ and morphisms the weakly monotone maps.

A \textbf{simplicial set} is a contravariant functor from $\Delta$ to sets, $X:\Delta^\op \to\Set$. The set of the $n$-simplices is denoted $X_n=X([n])$, for a morphism $\alpha:[n]\to[m]$ write $\alpha^*=X(\alpha):X_m\to X_n$.

The \textbf{singular complex} (singular simplicial set) of a space $Y$ is the simplicial set $$\S(Y):\Delta^\op \to\Set \qquad [n] \longmapsto \S(Y)_n :=\left\{\text{continuous maps }f:\nabla^n \to Y\right\}$$
where 
$$\nabla^n=\left\{(\xso)\in\R^{n+1}\middle| x_i\geq0,\ \sum x_i=1\right\}.$$

For $\alpha: [n]\to [m]$, the map $\alpha^*:\S(Y)_m\to\S(Y)_n$ is precomposition with the affine linear map $\alpha_*:\nabla^n\to\nabla^m$ defined by $e_i\mapsto e_{\alpha(i)}$ (i.e. the map $(\nno{t}{m})\mapsto\alpha_*(t)_j=\sum_{j=\alpha(i)}t_j$).

For a continuous map $\psi:Y\to Z$, a morphism of simplicial sets $\psi_*=\S(\psi):\S(Y)\to\S(Z)$ is given by $\S(\psi)_n(f)=\psi\circ f$. This yields a functor $\S:\Top\to\sSet$.

The \textbf{geometric realization} is a functor $|-|:\sSet\to\Top$ defined as follows. For a simplicial set $X$ we set
\[|X|=\left(\coprod X_n\times\nabla^n\right)/\sim\]
where $X_n$ is endowed with the discrete topology and the equivalence relation is the one generated by:\normalmarginpar\marginnote{\footnotesize The equivalence relation is only \textit{generated} by the condition stated, which is not symmetric. The actual equivalence relation is not easy to understand. We will return on this problem with the theory of minimal representatives in chapter \ref{subsection:minimal-representatives}.}
\[X_m\times\nabla^m\cont(x,\alpha_*(t))\sim(\alpha^*(x),t)\in X_n\times\nabla^n\quad\text{for all }\alpha:[n]\to[m],\ x\in X_m,\ t\in\nabla^n.\]

\reversemarginpar
