% Lecture 19

\chapter{Spaces and Simplicial Sets: an Introduction to Homotopy Theory (?)}\label{chapter:the-cool-chapter}

\Attention\ For the last part of the course, Professor Schwede decided to put me out of a job and upload \href{https://www.math.uni-bonn.de/people/schwede/sset_vs_spaces.pdf}{his own typed notes} (in English!). I will continue updating my notes as a way to study the material, but it goes without saying that the official version is more reliable.\medskip

\lecture[A lot of simplicial stuff, pt.2 (is this even topology anymore?).]{2021-12-20}

The long term goal for the rest of the semester is to show that in some sense (which we will specify), the \enquote{homotopy theories} of topological spaces and of simplicial sets are equivalent.

It is possible to prove (but it is beyond the scope of this course) that in fact there is an underlying equivalence of model categories/infinity categories; we will only see a \enquote{shadow} of this result, i.e. that there is an equivalence between the homotopy category of CW-complexes and the localization of the category of simplicial sets at the class of weak equivalences.\todo{Maybe I should rewrite this introduction}

The short term goal for this lecture and the next one is to show the existence of a preferred\rightnote{In this course, we do not like the word \enquote{canonical}.} CW-structure on $|X|$, the geometric realization of a simplicial set $X$.

\section{The Preferred CW-structure on a Geometric Realization}
Recall the \hyperref[section:reminder-on-sset]{basic theory of simplicial sets} as exposed in chapter I.

The geometric realization of a simplicial set is the space:
\[|X|=\left(\coprod X_n\times\ns\right)/\sim\]
where $X_n$ is endowed with the discrete topology and the equivalence relation is generated by:
\[X_m\times\nabla^m\cont(x,\alpha_*(t))\sim(\alpha^*(x),t)\in X_n\times\nabla^n\quad\text{for all }\alpha:[n]\to[m],\ x\in X_m,\ t\in\nabla^n.\]

Note that we did not give explicitly the equivalence relation $\sim$ on $|X|$, just a relation (not symmetric, in particular) that \tit{generates} it,
hence it might be difficult to tell when two points of $\coprod_{n\ge0} X_n\times\ns$ represent the same equivalence class in $|X|$.
To solve this issue, we need to study the equivalence relation $\sim$ more in detail.

\begin{remark*}
Given any relation $R$ on a set $X$, the equivalence relation generated by $R$ is the intersection of all the equivalence relations containing $R$, i.e. $a\sim b$ if and only if there exists $\xs\in X$ such that $a=x_0$, $b=x_n$ and $x_iRx_{i-1}$ or $x_{i-1}Rx_i$ for all $i=1,\dots,n$.
\end{remark*}

\subsection{Minimal Representatives}\label{subsection:minimal-representatives}

An important feature of the relation $\sim$ is that classes have minimal representatives, as shown by the following proposition.

\begin{proposition}[Minimal representatives]\label{proposition:minimal-representatives}
Let $X$ be a simplicial set.
\begin{numerate}
\item Every equivalence class for $\sim$ has a unique representative $(x,t)\in X_l\times\sx{l}$ of minimal dimension $l$, called the \tbf{minimal representative}.
\item A pair $(y,s)\in X_n\times\ns$ is the minimal representative in its class if and only if:
\begin{itemize}[label={-}]
    \item $y$ is a non-degenerate simplex,
    \item $s$ is an interior point of $\ns$.
\end{itemize}
\item If $(x,t)\in X_l\times\sx{l}$ is the minimal representative in its class and $(y,s)\in X_n\times\ns$ is equivalent to $(x,t)$, then there is a unique triple $(\delta,\sigma,u)$ consisting of:
\begin{itemize}[label={-}]
    \item an injective morphism $\delta:[k]\into[n]$,
    \item a surjective morphism $\sigma:[k]\onto[l]$,
    \item $u\in\ring\nabla^{k}$,
\end{itemize}
such that $\delta^*(y)=\sigma^*(x)$, $s=\delta_*(u)$ and $t=\sigma_*(u)$.
\end{numerate}
\end{proposition}

Summing up: the first part of the proposition gives existence of minimal representatives, the second part a characterization of them and the third says that any element in an equivalence class is related to the minimal representative of the class by a chain of just two \enquote{elementary equivalences} (the relations by which the equivalence relation $\sim$ is generated), i.e. $(y,s)=(y,\delta_*(u))\sim(\delta^*(y),u)=(\sigma^*(x),u)\sim(x,\sigma_*(u))=(x,t)$.

\begin{proof}
We write $X_l^\text{nd}$ for the set of non-degenerate $l$-simplices. We first define a map
\[\rho:\coprod_{n\ge0} X_n\times\ns\to\coprod_{l\ge0} X_l^\text{nd}\times\ring\nabla^l\]
such that $\rho(y,s)$ is equivalent to $(y,s)$.

Consider any $(y,s)\in X_n\times\ns$. Suppose that $s=(\nno{s}{n})$. Since $\nnso{s}{+}{n}=1$ and all $s_i\ge0$, there is at least a coordinate which is positive. Suppose that $k+1$ of the coordinates are positive. Define $u=(\nno{u}{k})$ to be the coordinates of $s$ with the zero entries deleted, in the same order. We let $\delta:[k]\to[n]$ be the unique injective morphism such that $\delta_*(u)=s$. The simplex $\delta^*(y)\in X_k$ can be written uniquely as $\delta^*(y)=\sigma^*(x)$ for a surjective morphism $\sigma:[k]\to[l]$ and a non-degenerate simplex $x\in X_l^\text{nd}$, by proposition \ref{proposition:non-degenerate-simplices}. Then we set:
\[\rho(y,s)=(x,\sigma_*(u)).\]
Since $u\in\ring\nabla^{k}$ and $\sigma_*:\nabla^{k}\to\nabla^{l}$ adds coordinates together (hence does not insert zeroes), also $\sigma_*(u)\in\ring\nabla^l$. Hence the map $\rho$ is well-defined and by construction $\rho(y,s)\sim(y,s)$.

Claim. If $(y,s)\in X_n\times\ns$ and $(\bar y,\bar s)\in X_{\bar n}\times\sx{\bar n}$ are equivalent pairs, then $\rho(y,s)=\rho(\bar y,\bar s)$.

\begin{claimproof}
It suffices to show this when $(y,s)$ and $(\bar y,\bar s)$ are "elementary equivalent", i.e. there is a morphism $\alpha:[n]\to[\bar n]$ such that
\[(y,s)=(\alpha^*(\bar y),s)\sim(\bar y,\alpha_*(s))=(\bar y,\bar s).\]
We let $(\delta,u,\sigma,x)$ be the data in the construction of $\rho(y,s)$ and we choose a factorization (which is necessarily unique) $\alpha\circ\delta=\bar\delta\circ\bar\sigma$ for a surjective morphism $\bar\sigma:[k]\to[\bar k]$ and an injective morphism $\bar\delta:[\bar k]\to[\bar n]$. Then
\[\bar s=\alpha_*(s)=\alpha_*(\delta_*(u))=\bar\delta_*(\bar\sigma_*(u)).\]

Since $u\in\ring\nabla^{k}$ and $\bar \sigma$ is surjective, $\bar\sigma_*(u)\in\ring\nabla^{\bar k}$. Hence $\bar s=\bar\delta_*(\bar\sigma_*(u))$ must be the unique expression in the first step of the construction of $\rho(\bar y,\bar s)$, since $\bar\delta_*$ is injective and $\bar\sigma_*(u)$ is an interior point. We write $\bar\delta^*(\bar y)=\hat\sigma^*(\hat x)$ for a surjective morphism $\hat\sigma:[\bar k]\to[\hat l]$ and a non-degenerate element $\hat x\in X_{\hat l}^\text{nd}$. Then we have:
\[\sigma^*(x)=\delta^*(y)=\delta^*(\alpha^*(\bar y))=\bar\sigma^*(\bar\delta^*(\bar y))=\bar\sigma^*(\hat\sigma^*(\hat x))=(\hat\sigma\circ\bar\sigma)^*(\hat x).\]
By the uniqueness of pairs of degeneracies and non-degenerate simplices (of proposition \ref{proposition:non-degenerate-simplices}), we must have $l=\hat l$, $x=\hat x$, $\sigma=\hat\sigma\circ\bar\sigma$. So $\bar\delta^*(\bar y)=\hat\sigma^*(\hat x)=\hat\sigma^*(x)$ and $(\bar\delta,\bar\sigma_*(u),\hat\sigma,x)$ must be the data in the construction of $\rho(\bar y,\bar s)$. Hence:
\[\rho(\bar y,\bar s)=(x,\hat\sigma_*(\bar\sigma_*(u)))=(x,\sigma_*(u))=\rho(y,s).\]
\end{claimproof}

(1) Suppose $(y,s)\in X_n\times\ns$ is of minimal dimension in its equivalence class. In the construction of $\rho(y,s)$ we must have $n\ge k\ge l$, therefore $n=k=l$, since $(y,s)\sim \rho(y,s)$. Clearly the injective map $\delta:[k]=[n]\into[n]$ and the surjective map $\sigma:[k]=[l]\onto[l]$ must be equal to $\id_{[n]}$. Hence $\rho(y,s)=(y,s)$.

If $(\bar y,\bar s)$ is another representative of the same minimal dimension, then
\[(y,s)=\rho(y,s)=\rho(\bar y,\bar s)=(\bar y,\bar s).\]
This also shows that $\rho(y,s)$ produces the unique minimal representative.

(2) The necessity falls out of the construction of $\rho$, which produces the minimal representative of a given class. Conversely, suppose $y$ is non-degenerate and $s$ an interior point. If $s=\delta_*(u)$ for another interior point $u$, then $\delta$ must be the identity, but then $\delta^*(y)=y=\sigma^*(x)$, hence also $\sigma$ must be the identity, so $(y,s)=\rho(y,s)$ is the minimal representative.

(3) Suppose $(x,t)$ minimal, and $(y,s)$ equivalent (but not elementary equivalent) to it. We want to find a unique triple $(\delta,\sigma,u)$ of an injective map, a surjective map and an interior point such that $\delta^*(y)=\sigma^*(x)$, $s=\delta_*(u)$ and $t=\sigma_*(u)$. The existence comes from the construction of $\rho(y,s)$, while the uniqueness is because no choices were involved (every step is uniquely determined).
\end{proof}

While the previous proof is not exactly enlightening, having minimal representatives makes it easier to work with the geometric realization functor.

\begin{corollary}\label{corollary:criterion-for-injectivity-of-map-between-realizations}
Let $f:X\to Y$ be a morphism of simplicial sets such that $f_n:X_n\to Y_n$ is injective for all $n\ge0$.
\begin{numerate}
\item For every non-degenerate $x\in X_n$, the simplex $f_n(x)\in Y_n$ is non-degenerate.
\item The continuous map $|f|:|X|\to|Y|$ is injective.
\end{numerate}
\end{corollary}

Note: if $f$ has a retract, then (2) follows by functoriality, however, having a retract is much stronger than having injective components (the left inverses of the components might not assemble into a natural transformation).

\begin{proof}\ 

(1) Suppose that $f_n(x)=s_i^*(y)$ for some $0\le i\le\ni$, $y\in Y_\ni$. Then (using that $f$ is a morphism of simplicial sets) we have:
\[f_n(s_i^*(d_i^*(x)))=s_i^*(d_i^*(f_n(x)))=s_i^*(\underbrace{d_i^*(s_i^*(y))}_{\id_{[n]}^*(y)})=s_i^*(y)=f_n(x).\]
Since $f_n$ is injective, $s_i^*(d_i^*(x))=x$, which contradicts the assumption that $x$ is non-degenerate.

(2) Let $(x,t)\in X_n\times\ns$ and $(x',t')\in X_m\times\sx{m}$ be minimal representatives of classes in $|X|$ such that $|f|[x,t]=|f|[x',t']$. By definition of the geometric realization functor we have $|f|[x,t]=[f_n(x),t]$ and $|f|[x',t']=[f_m(x'),t']$, and by part (1) both $f_n(x)$ and $f_m(x')$ are non-degenerate.
Then by uniqueness of minimal representatives we must have $(f_n(x),t)=(f_m(x'),t')$, hence in particular $n=m$, $t=t'$ and $x=x'$, i.e. $[x,t]=[x',t']$.
\end{proof}

\begin{corollary}\label{corollary:composition-of-realization}
Let $X$ be any simplicial set.
\begin{numerate}
\item The composite
\[\coprod_{n\ge0} X_n^\text{nd}\times\ns\into\coprod_{n\ge0} X_n\times\ns\onto|X|\]
is surjective.
\item The composite:
\[\coprod_{n\ge0} X_n^\text{nd}\times\ring\nabla^n\into\coprod_{n\ge0} X_n\times\ns\onto|X|\]
is a continuous bijection.
\end{numerate}
\end{corollary}

We will see that the second statement gives us the decomposition into cells of the preferred CW-structure for the geometric realization of a simplicial set.

We also have a useful corollary of (1).\rightnote{Note: $|X|$ will turn out to be Hausdorff once we show it is a CW-complex, so we do not try to prove it directly (although it should be feasible).}

\begin{corollary}\label{corollary:criterion-for-compactness-of-realization}
Suppose that the total number of non-degenerate simplices in all dimensions is finite. Then $|X|$ is quasi-compact.
\end{corollary}

\subsection{The Skeleta filtration}

The preferred CW-structure of $|X|$ arises from a \enquote{CW-like structure} that is intrinsic to the combinatorics of simplicial sets.

The \tbf{$m$-skeleton} $\sk^m X$ of a simplicial set $X$, for $m\ge0$, is the simplicial set defined by:
\[(\sk^m X)_n=\cb{x\in X_n\mid x=\alpha^*(y)\text{ for some }\alpha:[n]\to[m]\text{ and some }y\in X_m},\]
i.e. $\sk^m X$ is the smallest simplicial subset that contains $X_m$.

Clearly one has to convince himself that this is a simplicial set, which is not difficult.

\begin{example}
Every constant simplicial set (i.e. constant as a presheaf) is zero-dimensional, i.e. $X=\sk^0 X$. Conversely if $X=\sk^0 X$, then $X$ is isomorphic to the constant simplicial set on the zero-simplices $X_0$.
\end{example}

\begin{example}
For $\Delta^m=\Delta(-,[m])$, the simplicial $m$-simplex, we have $\sk^m(\Delta^m)=\Delta^m$ and $\sk^{m-1}(\Delta^m)=\de\Delta^m$, where $(\de\Delta^m)_n=\cb{\alpha:[n]\to[m]\mid\alpha\text{ not surjective}}$.
\end{example}

Note: these examples may not be necessarily obvious at first, one should diligently sit down and check the details.
